Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-13 vertical-thermal-review.Rmd
Tleaf section

**The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher wind speeds (e.g., ~ 4 m s^-1^; Fig. `r fig_NEON_vertical`) can reduce $T_{leaf}-T_{air}$ to ~ $4^\circ$C [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Collectively, higher wind speed (Fig. `r fig_NEON_vertical`), and with adaptive leaf traits increase $g_b$ and $g_s$ (reviewed below) mediate the direct effects of solar radiation on upper canopy leaves (Fig. `r fig_leaf_T`a) and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling. Under lower wind conditions, however,  lower $g_b$ increases $T_{leaf} - T_{air}$ [Fig. `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat or drought stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].

>>>>

The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf metabolism and thermal responses" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions. Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$-limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].
